---
title: "Data Analysis and Visualisation"
subtitle: "Sessions 4 & 5"
author: "Thomas Wise"
format: 
  revealjs:
    smaller: true
    theme: dark
editor: visual
---

## Session Objectives (Session 4)

- Understand how to conduct simple analysis 
- Understand how to use the `apply` family of functions 
- Understand how to apply Foundational statistical tests

## Today's Plan (Session 4)

- Initial Discussion
- Section 0: Session Setup
- Section 1: Conducting Simple Analysis 
  - Part 1.1: Applying analytic functions 
  - Part 1.2: Using the `apply` family of functions 
- Section 2: Conducting Foundational Statistical Tests 
  - Part 2.1: Using Correlations 
  - Part 2.2: Using Regressions 
  - Part 2.3: Other Statistical Tests 

## Initial Discussion 

- Do you have any questions or follow up from last weeks session? 

- Are there any statistical tests, analysis techniques you would like to cover today or next week? 
- Are there any visualisation you'd like to learn how to create? 

## Section 0: Session Setup

- Was everyone able to access the data?
- Can everyone access the materials, RStudio etc?
- Can everyone open up the script labelled *worksheet_dataanalysis_session4.R*?

# Section 1: Conducting Simple Analysis

## Part 1.1: Applying analytic functions 

In previous sessions we have already covered one method:

- `summary()`

But sometimes you may want to get specific information for calculations:

- Minimum and Maximum: `min()` and `max()`
- Mean - `mean()`
- Median - `median()`
- Standard Deviation - `sd()`
- Quantile - `quantile()`

## Part 1.1: Applying analytic functions 

You can compile you own grouped statistics, through using the `summarise()` function

- This allows you to utilise the `group_by()` function, before creating a descriptive statistic table from scratch.

```{r}
#| echo: TRUE
#| eval: FALSE

# Example 

data %>% 
  group_by(val) %>% 
  summarise(mean = mean(x),
            min = min(x),
            max = max(x))
```

## Part 1.2: Using the `apply` family of functions

`apply()` is a family of functions which allows functions to be applied to the rows or columns of a data frame / matrix

This includes:

| Function   | Use on                    | Parameters             |
|------------|---------------------------|------------------------|
| `apply()`  | matrix or dataframe       | x, margin, function    |
| `lapply()` | list                      | x, function            |
| `sapply()` | list, vector or dataframe | x, function            |
| `tapply()` | dataframe                 | x, index, function     |
| `mapply()` | multiple lists or vectors | listX, listY, function |

# Section 2: Conducting Foundational Statistical Tests 

## Part 2.1: Using Correlations 

Two types of functions:

- `cor(x, y, method)` for the correlation coefficient
- `cor.test(x, y, method)` as a test of association between a paired sample

Method types:

- `pearson`
- `kendall`
- `spearman`

## Part 2.2: Using Regressions 

Simplest regression function in R is:

- `lm(formula, data)`
- `glm(formula, family = gaussian, data)`

More complex functions exist, but this moves into Machine Learning territory!

- Check out `glmnet()`

## Part 2.2: Interpretation of Regressions 

Key interpretations:

- *Coefficients*: Provide information about the variables, such as significance\
- *Residual Standard Error*: Used as a measure of fit between model and data set
- *Multiple/Adjusted R-Squared*: How well does the model explain variance in the observed data
- *F-statistic*: Used to test the significance of coefficients
- *p-value*: NHST significance level

## Part 2.3: Other Statistical Tests 

- t-tests: `t.test()`
- ANOVA: `aov()`
- MANOVA: `manova()`
- Wilcox-test: `wilcox.test()`
- Chi-square test: `chisq.test()`

And Many many more!! 

## Resources and Further Reading 

- [R for Data Science, Wickham, Cetinkaya-Rundel and Grolemund](<https://r4ds.hadley.nz/>)
- [Modern Statistics with R, Mans Thulin](<https://www.modernstatisticswithr.com/>)
- [Learning Statistics with R - Danielle Navarro](https://learningstatisticswithr.com/lsr-0.6.pdf)
- [Introduction to Statistical Learning - James, Witten, Hastie & Tibshirani](https://www.statlearning.com/)

# Session 5

## Session Objectives (Session 5)

- Understand some foundational principles of data visualisation
- Understand the syntax of visualisations using `ggplot2`
- Understand how to create visualisations using `ggplot2`

## Today's Plan (Session 5)

- Initial Discussion
- Section 0: Session Setup
- Section 1: Foundational Data Visualisation Theory  
- Part 1.1: Visualisation Best Practice 
- Section 2: Using `ggplot2` to create visualisations  
- Part 2.1: `ggplot2` syntax 
- Part 2.2: Developing Visualisations  

## Data Visualisation - Reference Materials / Guidance

- [ggplot2 Tidyverse overview](https://ggplot2.tidyverse.org/)

- [ggplot2 - Full Literature, 3rd Edition](https://ggplot2-book.org/)

- [R4DS - Section on Data Visualisation](https://r4ds.had.co.nz/data-visualisation.html)

## Section 0: Session Setup

- Was everyone able to access the data?
- Can everyone access the materials, RStudio etc?
- Can everyone open up the script labelled *worksheet_datavisualisation_session4.R*?

# Section 1: Foundational Data Visualisation Theory 

## Part 1.1: Visualisation Best Practice

The analysis function published guidance on accessible plots:

- [Data Visualisation: Charts](https://analysisfunction.civilservice.gov.uk/policy-store/data-visualisation-charts/)
- [Making analytical Publications accessible](https://analysisfunction.civilservice.gov.uk/policy-store/making-analytical-publications-accessible/#section-4)

These guidelines are complemented by and also further developed by:

- Gestalt Theory
- E. Tufte Best Practice

## Part 1.1: Analytical Function Guidance

As a public sector service, we must meet Accessibility Legislation guidance (2020).

Although this guidance is aimed towards publications rather than internal documentation, following these internally is a good practice and reduces your later processing time to ensure it meets accessibility standards.

Throughout today's session we will refer back to Analytical functions guidance, so help guide the session

## Part 1.1: Gestalt Theory

Emerging from a school of psychology, Gestalt theory states that we perceive objects as a whole, rather than a sum of its parts.

This field highlights many cognitive shortcuts we can use in our data visualizations

- Enclosure: Boundaries demonstrate grouping
- Similarity: Items which share features (colour, shape, size) are seen as a group
- Continuity: Aligned or flowing objects are seen as a group
- Connection: Connected objects, regardless of attribute are seen as grouped
- Closure: Incomplete boundaries are auto-completed
- Proximity: Objects close together are perceived as similar
- Symmetry: Symmetric objects are pleasing, asymetric ones stand out

## Part 1.1: E. Tufte Best Practice

E. Tufte, American Statistician, Political Scientist and Computer Science Emeritus Professor highlights 6 core principles of best practice:

- Comparisons: Show data by comparisons, depict contrasts and differences
- Causality: Demonstrate the influence or impact of variables
- Multivariate: Combine data where appropriate to help interpret complex narratives
- Integration: Incorporate various modes of information to further highlight evidence
- Documentation: For credibility, include attribution, detailed titles and measurements
- Context: Describe or depict the before and after state

## Part 1.1: E. Tufte Best Practice

These can be applied through the following methods:

- Graphical Integrity
  - Be sure that the graph is telling the truth and is representative of such
- Data-Ink Ratio
  - Maximize the data-ink ratio, and erasing redundant data/non-data ink
- Limiting Chart Junk
  - Removing or limiting non-data specific "junk"
- Data Density
  - Have an appropriate density of data and information in your graphic space
- Small Multiples
  - Use compressed, matching, small style graphs to visualise large quantities of information

# Section 2: Using `ggplot()` to create visualisations

## Part 2.1: `ggplot()` syntax: Anatomy 101

Fundamental Elements:

- `ggplot()` - specifies that a graph will be produced
- `geom_XX()` - specifies the type of graph

Key Parameters:

- `data =` - specify your data
- `mapping = aes(x, y)` - specify what you'd like to map onto the graph
- `x`, `y` - specify the x/y axis respectively
- `colour` - specify the colour
- `alpha` - specify the transparency
- `size` - specify the size of the points

## Part 2.1: Example Plot

```{r}
#| warnings: FALSE
#| include: FALSE
library(tidyverse)

```

```{r}
#| echo: TRUE

# Example 

ggplot(data = diamonds) + 
  geom_point(mapping = aes(x = carat, y = price))
```

## Part 2.1: Common Visualisation Types 

- Scatter plots: `geom_point()`
- Jittered Scatter plots: `geom_jitter()`
- Bar Charts: `geom_col()` or `geom_bar()`
- Histogram: `geom_histogram()`
- Density plot: `geom_density()`

## Part 2.2: Visualisations are like Onions....

- Visualisations can be built up, using layers

- Each layer is added in order, meaning it is possible to build up a visualisation

- This can be achieved by adding `+`, and adding another layer of `geom_()` as they are standardised.

## Part 2.2: .... they have layers

However:

- Each layer automatically inherits the information provided in the initial `ggplot()` call, so be sure to specify information by layer if unique and not shared
- Each layer is added in order, so setting `alpha` (transparency) is important!

## Part 2.2: Specifying Axis's

Simple shortcuts:

- `lims(x = c(min, max), y = c(min, max))`

Advanced used:

- `scale_x_continuous()` / `scale_y_continuous()`
- `scale_x_sqrt()` / `scale_y_sqrt()`
- `scale_x_log10()` / `scale_y_log10()`

With these having the internal arguments:

- breaks, minor_breaks, n.breaks
- labels, limits

## Part 2.2: Specifying Labels

You can simply specify axis labels using the `labs()` function with the internal arguments:

- `title =` for the plot title
- `subtitle =` for the plot subtitle
- `caption =` for the plot caption
- `x` / `y` for x or y axises respectively

Advanced labels can be given through:

- `annotate()` function combined with geom `text` or `rect`
- `geom_vline()` for a vertical line
- `geom_hline()` for a horizontal line

## Part 2.2: Using preset themes

- `theme_bw()`
- `theme_classic()`
- `theme_dark()`
- `theme_light()`
- `theme_minimal()`
- `theme_void()`

## Part 2.2: Using specifying manual themes

By using the `theme()` function as a foundation, you can specify almost every element of a visualisation plot.

But remember to call these functions when using specific elements

- Line elements: `element_line()`
- Rectangular/area elements: `element_rect()`
- Text elements: `element_text()`

You can use these to rotate axis labels too! For example:

- `theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))`

# Reflections 

# Feedback

